Designing a high-performance boundary element library with OpenCL and Numba

### 2021-04-06 bem.tex
american spelling

\section{Boundary element methods with Bempp}
In this section, we provide a brief introduction to boundary element methods and describe the necessary steps for their numerical discretisation and solution.

@@

As a concrete example, we consider computing the electrostatic capacity of an object $\Omega$. In this case, we solve the above equation with $f(\bx)=1$ and $g(\bx,\by)=\frac{1}{4\uppi|\bx - \by|}$. Once $\phi$ has been found, the normalised capacity is then obtained using $c = \frac{1}{4\uppi}\int_{\Gamma}\phi(\bx)\ds[\bx]$.

@@

\begin{figure}
  \centering
  \input{img/sphere}
  \caption{Discretisation of a sphere into flat surface triangles.}
  \label{fig:triangulation}
\end{figure}

The first step is the \textbf{discretisation of the surface} $\Gamma$. Surfaces are represented in Bempp-cl as a triangulation into flat triangles (see \cref{fig:triangulation}). The triangulation is internally represented as an array of node coordinates and an associated connectivity array of node indices that define each triangle. In this step, topology data is also computed: in particular, for each triangle, we compute the neighboring triangles and the type of intersection (i.e. are they connected by an edge or a vertex).

Once a triangulation is given we need to define the necessary data structures for the discretisation. Bempp-cl uses a Galerkin discretisation: we introduce set of basis functions \(\phi_1\) to \(\phi_N\), and define the \textbf{trial space} as the span of these function. We then approximate the solution \(\phi\) of \cref{eq:bnd_integral} by $\phi_h=\sum_{j=1}^N \dvec{x}_j\phi_j$, where $\dvec{x}$ is a vector of coefficients. In the most simple case, we can define the function \(\phi_j\) to be equal to 1 on the triangle $\tau_j$ and 0 everywhere else. Other spaces are commonly defined to be piecewise polynomials on each triangle.

To discretise \cref{eq:bnd_integral}, we define a \textbf{test (or dual) space} in terms of a basis \(\psi_1\) to \(\psi_N\). The discrete representation of the above problem then takes the form
$$
\dmat{A}\dvec{x} = \dvec{b}
$$
with
\begin{align*}
\dmat{A}_{ij} &= \int_{\Gamma}\psi_i(\bx)\int_{\Gamma}g(\bx, \by)\phi_j(\bx)\ds[\by]\ds[\bx]\\
\dvec{b}_i &= \int_{\Gamma}\psi_i(\bx)f(\bx)\ds[\bx].
\end{align*}
In the case of piecewise constant trial and test functions, the definition of $\dmat{A}_{ij}$ simplifies to $\dmat{A}_{ij} = \int_{\Gamma}\int_{\Gamma}g(\bx, \by)\ds[\by]\ds[\bx]$.

@@

In the final step, \textbf{Bempp solves the underlying linear system of equations} either through a direct LU decomposition or through iterative solvers such as GMRES. The solution can then be evaluated away from the surface $\Gamma$ through domain potential operators and exported in various formats for visualisation.

In summary, to solve a boundary integral equation problem, the following steps are performed by Bempp:
\begin{enumerate}
	\item Import of the surface description as triangulation data.
	\item Definition of the trial space, test space and range space.
	\item Discretisation into a matrix problem $\dmat{A}\dvec{x}=\dvec{b}$.
	\item Solution of the matrix problem by either a direct or iterative solver.
	\item Evaluation of domain potential operators for visualisation and post-processing.
\end{enumerate}

>>>>

\section{Boundary element methods with Bempp}
In this section, we provide a brief introduction to boundary element methods and describe the necessary steps for their numerical discretization and solution.

@@

As a concrete example, we consider computing the electrostatic capacity of an object $\Omega$. In this case, we solve the above equation with $f(\bx)=1$ and $g(\bx,\by)=\frac{1}{4\uppi|\bx - \by|}$. Once $\phi$ has been found, the normalized capacity is then obtained using $c = \frac{1}{4\uppi}\int_{\Gamma}\phi(\bx)\ds[\bx]$.

@@

\begin{figure}
  \centering
  \input{img/sphere}
  \caption{Discretization of a sphere into flat surface triangles.}
  \label{fig:triangulation}
\end{figure}

The first step is the \textbf{discretization of the surface} $\Gamma$. Surfaces are represented in Bempp-cl as a triangulation into flat triangles (see \cref{fig:triangulation}). The triangulation is internally represented as an array of node coordinates and an associated connectivity array of node indices that define each triangle. In this step, topology data is also computed: in particular, for each triangle, we compute the neighboring triangles and the type of intersection (i.e. are they connected by an edge or a vertex).

Once a triangulation is given we need to define the necessary data structures for the discretization. Bempp-cl uses a Galerkin discretization: we introduce set of basis functions \(\phi_1\) to \(\phi_N\), and define the \textbf{trial space} as the span of these function. We then approximate the solution \(\phi\) of \cref{eq:bnd_integral} by $\phi_h=\sum_{j=1}^N \dvec{x}_j\phi_j$, where $\dvec{x}$ is a vector of coefficients. In the most simple case, we can define the function \(\phi_j\) to be equal to 1 on the triangle $\tau_j$ and 0 everywhere else. Other spaces are commonly defined to be piecewise polynomials on each triangle.

To discretize \cref{eq:bnd_integral}, we define a \textbf{test (or dual) space} in terms of a basis \(\psi_1\) to \(\psi_N\). The discrete representation of the above problem then takes the form
$$
\dmat{A}\dvec{x} = \dvec{b}
$$
with
\begin{align*}
\dmat{A}_{ij} &= \int_{\Gamma}\psi_i(\bx)\int_{\Gamma}g(\bx, \by)\phi_j(\bx)\ds[\by]\ds[\bx]\\
\dvec{b}_i &= \int_{\Gamma}\psi_i(\bx)f(\bx)\ds[\bx].
\end{align*}
In the case of piecewise constant trial and test functions, the definition of $\dmat{A}_{ij}$ simplifies to $\dmat{A}_{ij} = \int_{\Gamma}\int_{\Gamma}g(\bx, \by)\ds[\by]\ds[\bx]$.

@@

In the final step, \textbf{Bempp solves the underlying linear system of equations} either through a direct LU decomposition or through iterative solvers such as GMRES. The solution can then be evaluated away from the surface $\Gamma$ through domain potential operators and exported in various formats for visualization.

In summary, to solve a boundary integral equation problem, the following steps are performed by Bempp:
\begin{enumerate}
	\item Import of the surface description as triangulation data.
	\item Definition of the trial space, test space and range space.
	\item Discretization into a matrix problem $\dmat{A}\dvec{x}=\dvec{b}$.
	\item Solution of the matrix problem by either a direct or iterative solver.
	\item Evaluation of domain potential operators for visualization and post-processing.
\end{enumerate}

### 2021-04-06 bempp_overview.tex
american spelling

Once a user has defined in operator using Bempp-cl, the discretisation can be computed by calling the \pyth{weak_form} method. Upon calling this method, a regular integrator will be used to assemble all the interactions between non-adjacent elements, and a singular integrator will be used to compute the interactions between adjacent triangles (if the trial and test spaces are defined on different grids, this second integrator is not needed). Depending on the user's preferences, these integrators will internally use computational routines defined using either Numba or OpenCL.

For OpenCL assembly, the code checks additional parameters, such as the default vector length for SIMD operators (e.g., 4 for double precision and 8 for single precision in Intel AVX2, or 1 if a GPU is used), andwhether the discretisation should proceed in single or double precision. The OpenCL kernel is then compiled for the underlying compute device using PyOpenCL and executed. If the computational backend is Numba, the call is forwarded to the corresponding Numba discretisation routines and executed.

For simple piecewise constant function spaces or other spaces, where the support of each basis function is localised to a single triangle, only one call to the computational routines is necessary. If the support of basis functions is larger than a single triangle, different threads may need to sum into the same global degree of freedom. We discuss this in more detail in \cref{sec:coloring}.

Outside the operator discretisation, Numba is used in the following contexts:
\begin{itemize}
  \item Computing the grid topology: this involves iterating through the grid to compute the neighbour relataionships between triangles.
  \item Definition of local-to-global maps for function spaces: again, this requires traversal through the grid and assigning relationships between global and local indices.
  \item Grid functions: a right-hand side function $f$ can be defined as a Python callable. This is just-in-time compiled via Numba and then the product with the corresponding basis functions is integrated in each triangle via numerical quadrature, again via Numba accelerated routines.
  \item Computing sparse matrices.
\end{itemize}

As the cost of each of these processes is in general much smaller than the cost of operator discretisation, these can be permformed using Numba without any need to consider the use of OpenCL for potential further speed up.

>>>>

Once a user has defined in operator using Bempp-cl, the discretization can be computed by calling the \pyth{weak_form} method. Upon calling this method, a regular integrator will be used to assemble all the interactions between non-adjacent elements, and a singular integrator will be used to compute the interactions between adjacent triangles (if the trial and test spaces are defined on different grids, this second integrator is not needed). Depending on the user's preferences, these integrators will internally use computational routines defined using either Numba or OpenCL.

For OpenCL assembly, the code checks additional parameters, such as the default vector length for SIMD operators (e.g., 4 for double precision and 8 for single precision in Intel AVX2, or 1 if a GPU is used), andwhether the discretization should proceed in single or double precision. The OpenCL kernel is then compiled for the underlying compute device using PyOpenCL and executed. If the computational backend is Numba, the call is forwarded to the corresponding Numba discretization routines and executed.

For simple piecewise constant function spaces or other spaces, where the support of each basis function is localized to a single triangle, only one call to the computational routines is necessary. If the support of basis functions is larger than a single triangle, different threads may need to sum into the same global degree of freedom. We discuss this in more detail in \cref{sec:coloring}.

Outside the operator discretization, Numba is used in the following contexts:
\begin{itemize}
  \item Computing the grid topology: this involves iterating through the grid to compute the neighbour relataionships between triangles.
  \item Definition of local-to-global maps for function spaces: again, this requires traversal through the grid and assigning relationships between global and local indices.
  \item Grid functions: a right-hand side function $f$ can be defined as a Python callable. This is just-in-time compiled via Numba and then the product with the corresponding basis functions is integrated in each triangle via numerical quadrature, again via Numba accelerated routines.
  \item Computing sparse matrices.
\end{itemize}

As the cost of each of these processes is in general much smaller than the cost of operator discretization, these can be permformed using Numba without any need to consider the use of OpenCL for potential further speed up.

### 2021-04-06 intro.tex
american spelling

This proved a burden when efforts started to modernise the library to be able to make better use of single-instruction-multiple-data (SIMD) optimisation on CPUs and offload computation to GPUs. In 2018, we decided to rewrite the computational core from scratch. The aims were to support explicit SIMD optimisation on CPUs with various instruction lengths, be able to offload computations to AMD, Intel, and Nvidia GPUs, and to base the complete codebase on Python. These aims naturally led to the choice of building a Python library based around OpenCL (using the PyOpenCL interface) and Numba.

>>>>

This proved a burden when efforts started to modernize the library to be able to make better use of single-instruction-multiple-data (SIMD) optimization on CPUs and offload computation to GPUs. In 2018, we decided to rewrite the computational core from scratch. The aims were to support explicit SIMD optimization on CPUs with various instruction lengths, be able to offload computations to AMD, Intel, and Nvidia GPUs, and to base the complete codebase on Python. These aims naturally led to the choice of building a Python library based around OpenCL (using the PyOpenCL interface) and Numba.
